\chapter{Third-Party Code and Libraries}

\section{JavaScript A* Pathfinding Library}\label{pathfinding_library}

This library is used to supply A* pathfinding for movement in the game's client\cite{citeulike:13160393}. The library is released under the MIT License
\cite{MITLicense} and is used with some modification.

\section{Flask Microframework}

This framework supplies the routing and HTML template rendering functionality, as well as the base for other plugins that are used on the server\cite{citeulike:13160396}. It is distributed under the BSD License \cite{FlaskBSDLicense} and is used without modification.

\subsection{Flask-SocketIO}
A plugin for the Flask framework implementing the SocketIO library. It is distributed under the MIT License\cite{MITLicense} and is used without modification.

\subsection{Flask-SQLAlchemy}
A plugin for the Flask framework implementing the Python SQLAlchemy ORM. It is distributed under Flask's BSD License \cite{FlaskBSDLicense} and is used without modification.

\subsection{Flask-WTF}
A plugin for the Flask framework implementing the Python WTForms library for managing web forms. It is distributed under a BSD License \cite{WTFBSDLicense} and is used without modification.

% If you have made use of any third party code or software libraries, i.e. any code that you have not designed and written yourself, then you must include this appendix.

% As has been said in lectures, it is acceptable and likely that you will make use of third-party code and software libraries. The key requirement is that we understand what is your original work and what work is based on that of other people.

% Therefore, you need to clearly state what you have used and where the original material can be found. Also, if you have made any changes to the original versions, you must explain what you have changed.

% As an example, you might include a definition such as:

% Apache POI library – The project has been used to read and write Microsoft Excel files (XLS) as part of the interaction with the client’s existing system for processing data. Version 3.10-FINAL was used. The library is open source and it is available from the Apache Software Foundation
% \cite{apache_poi}. The library is released using the Apache License
% \cite{apache_license}. This library was used without modification.
