\chapter{Original Proposal}\label{appendix:designspec}

An online, multiplayer roleplaying game similar to \textit{Dungeons \& Dragons} hosted in the browser. The game will have a Game Master and several players in a group. The players move around the world and interact with it and the people/creatures in it and the Game Master guides the players, sets rules and manages the experience. They will be able to manipulate the game world and do things players can't do. For example, they could teleport a player to another location, spawn new monsters or change the weather.

It will include isometric graphics and aim to be as permissive as possible in what players are allowed to do (for example, interact with the environment using items they have, such as burning down a house with a torch). This means that there will be several systems, including combat, movement, character/creature creation, interacting with other characters (players and non-players, with chat as well as game system rules) and the environment (things like objects having heat and a burning point, in the fire example).

The nature of the design means that the project is open-ended, allowing for a basic game with a few simple systems and a single dungeon and character type to a game with several systems, full environmental interaction, a map and character editor and the ability to upload your own artwork.

It will require a server to manage the clients and let them interact, relay chat and act as an authority in the game world to keep clients in sync. It will also have to keep track of where players are in the world and whether they can see and interact with each other. There will be a graphical front end client that the players use and, in theory, multiple different clients could be made, such as a mobile app.